Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2022-01-05 analysis/paper/London_HauloutBeahvior.Rmd
CHANGE CAPTION & OTHER SMALL BITS

# bar chart instead of a strip
#
# ggplot(data = seal_hours_tbl, aes(x=yday, y=n, fill=n)) +
#   geom_col(binwidth = 1) +
#   scale_fill_gradient(low = "#6baed6", high = "#08306b") +
#   scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
#                      labels = c("March", "April", "May", "June", "July"),
#                      expand = expansion(mult = c(0.01, 0.025))) +
#   guides(fill = guide_legend(title.position = "bottom", title.hjust = 0.5)) +
#   facet_grid(species ~ .) + 
#   theme(axis.text.x = element_text(hjust = 0),
#         axis.text.y = element_blank(),
#         axis.title = element_blank(),
#         panel.grid = element_blank(),
#         legend.position = "bottom") +
#   labs(fill = "seal-hours of bio-logger data") +
#   ggtitle("Density of haul-out behavior observations varies from March through July",
#           subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")

@@

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
[@verhoef2010a] differ between models [@teneyck2018]. After fitting GLMPM
models, we instead used "type III" $F$-tests to calculate $p$-values
[@verhoef2010a] to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of influential
predictors (e.g. solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces, explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea ice extent
in the study region.

Visualizing the marginal or conditional effect of an individual weather
covariate on haul-out probability is difficult in this analysis because of the
collinearity between covariates as well as the spatial and temporal variation
across such a large region. The relationship of each weather covariate with
haul-out probability, averaged over the other weather conditions, is more
variable than simply showing model coefficients would imply. To elucidate
underlying patterns between the model's predicted haul-out probability and
specific weather covariates, we explored a subset of predictions during three
separate time windows. The same set of predictions used to create the haul-out
probability surfaces was reduced to include only local solar hours between 06:00
and 18:00 during three 7-day time windows. The time windows were centered on the
day of maximum haul-out probability and 25 days before and 25 days after peak
haul out. While this approach does not entirely remove patterns of collinearity
from the visualization (e.g. temperature increases with day of year), by
limiting our range of days and hours we gain some insight into how haul-out
probability varies across different weather covariate values. Figures
\@ref(fig:beardedHOwx), @\ref(fig:ribbonHOwx), and @\ref(fig:spottedHOwx)
present the predicted haul-out probability with respect to age and sex class
across the range of temperature and wind conditions encountered in the three
time windows. The visualizations also include vertical lines representing 95%
confidence intervals around the predicted haul-out probability to better
communicate the variation in model uncertainty. Additionally, simple linear fits
are shown to assist interpretation of overall patterns.

@@

@@

p_temp / p_wind +
  plot_annotation(
    title = 'The influence of temperature and wind on predicted bearded seal \nhaul-out behavior varies within season',
    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate marginal patterns.')
```

@@

p_temp / p_wind +
  plot_annotation(
    title = 'The influence of temperature and wind on predicted ribbon seal \nhaul-out behavior varies within season',
    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate marginal patterns.')
```

@@

p_temp / p_wind +
  plot_annotation(
    title = 'The influence of temperature and wind on predicted spotted seal \nhaul-out behavior varies within season',
    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate marginal patterns.')
```

>>>>

@@

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
[@verhoef2010a] differ between models [@teneyck2018]. After fitting GLMPM
models, we instead used "type III" $F$-tests to calculate $p$-values
[@verhoef2010a] to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of three influential
predictors (e.g. solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces (see figures \@ref(beardedHOCal), \@ref(ribbonHOCal), and
\@ref(spottedHOCal)), explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea ice extent
in the study region.

Visualizing the marginal or conditional effect of an individual weather
covariate on haul-out probability is difficult in this analysis because of the
collinearity between covariates as well as the spatial and temporal variation
across such a large region. The relationship of each weather covariate with
haul-out probability, averaged over the other weather conditions, is more
variable than simply showing model coefficients would imply. To elucidate
underlying patterns between the model's predicted haul-out probability and
specific weather covariates, we explored a subset of predictions during three
separate time windows. The same set of predictions used to create the haul-out
probability surfaces was reduced to include only local solar hours between 06:00
and 18:00 during three 7-day time windows. The time windows were centered on the
day of maximum haul-out probability and 25 days before and 25 days after peak
haul out. While this approach does not entirely remove patterns of collinearity
from the visualization (e.g. temperature increases with day of year), by
limiting our range of days and hours we gain some insight into how haul-out
probability varies across different weather covariate values. Figures
\@ref(fig:beardedHOwx), \@ref(fig:ribbonHOwx), and \@ref(fig:spottedHOwx)
present the predicted haul-out probability with respect to age and sex class
across the range of temperature and wind conditions encountered in the three
time windows. The visualizations also include vertical lines representing 95%
confidence intervals around the predicted haul-out probability to better
communicate the variation in model uncertainty. Additionally, simple linear fits
are shown to assist interpretation of overall patterns.

@@

p_wind <- ggplot(plot_df, aes(wind*10, ho_prob, color = age_sex, fill = age_sex)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.1, stroke = 0) +
  geom_smooth(method = "lm", formula = y ~ x, size = 0.3, alpha = 0.1,
              show.legend = FALSE) +
  scale_fill_d3() +
  scale_color_d3(guide = guide_legend(override.aes = list(alpha = 0.75),
                                      direction = "horizontal")) +
  facet_wrap(. ~ season + date_start + date_end, scales = "free_x",
             labeller = label_glue('{season}\n{date_start} - {date_end}')) +
  theme(legend.title = element_blank(),
        legend.position = "none") +
  xlab("total wind (m/s)") + ylab("haul-out probability")

@@

p_temp / p_wind +
  plot_annotation(
    title = 'The influence of temperature and wind on predicted bearded seal \nhaul-out behavior varies within season',
    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate underlying trends.')
```

@@

p_temp / p_wind +
  plot_annotation(
    title = 'The influence of temperature and wind on predicted ribbon seal \nhaul-out behavior varies within season',
    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate underlying trends.')
```

@@

p_temp / p_wind +
  plot_annotation(
    title = 'The influence of temperature and wind on predicted spotted seal \nhaul-out behavior varies within season',
    subtitle = 'data are subset to only include local solar hours between 0800 and 1600',
    caption = 'Solid lines represent linear fits to the predicted values and are \nprovided to better indicate underlying trends.')
```
